---
title: "Experimental Designs and Estimators"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Experimental Designs and Estimators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(CausalModel)
set.seed(2024)
```

## Overview

This vignette mirrors the Python `experimental.ipynb` notebook and shows the
design objects and experimental estimators in CausalModel.

## Design Objects

CausalModel has two randomization designs: completely randomized design (CRD)
and Bernoulli randomization.

### Completely Randomized Design

```{r crd-basic}
d <- crd()
draw(d, 10)
```

### CRD with Covariate Balance

Rerandomization to ensure covariate balance (Morgan & Rubin, 2012).
The design draws assignments until Mahalanobis balance falls below a threshold:

```{r crd-balance}
X <- matrix(rnorm(100 * 2), ncol = 2)
d_bal <- crd(covariate = X, balance = TRUE)
Z_bal <- draw(d_bal, 100)
cat("Treatment assigned:", sum(Z_bal), "of", length(Z_bal), "\n")
```

### Bernoulli Randomization

Each unit is independently assigned to treatment with a fixed probability:

```{r bernoulli}
d_bnli <- bernoulli()
draw(d_bnli, 10)
```

## Experimental Estimators

### Setup

Generate experimental data with a known treatment effect $\tau = 10$:

```{r exp-data}
data <- generate_data(N = 10000, k = 2, tau = 10)
exp_obj <- experimental(data$Y, data$Z, data$X)
```

### Difference-in-Means

The default estimator under a CRD:

```{r dm}
est_via_dm(exp_obj)
```

### Stratified Estimator

Weighted average of within-stratum difference-in-means. Strata are defined
by the user — here we use a simple median split on the first covariate:

```{r strata}
strata <- as.integer(data$X[, 1] > median(data$X[, 1]))
est_via_strata(exp_obj, strata = strata)
```

With random strata (less informative):

```{r strata-random}
strata_rand <- sample(0:2, length(data$Y), replace = TRUE)
est_via_strata(exp_obj, strata = strata_rand)
```

### ANCOVA

Adjusts for covariates via $Y \sim 1 + Z + X + Z \cdot X$ with HC0 robust
standard errors. Achieves efficiency gains when covariates are predictive:

```{r ancova}
est_via_ancova(exp_obj)
```

### Fisher Randomization Test

Tests the sharp null hypothesis of zero individual treatment effects by
comparing the observed test statistic to its randomization distribution:

```{r fisher}
pval <- test_via_fisher(exp_obj, n_draws = 1000)
cat("Fisher p-value:", pval, "\n")
```

With $\tau = 10$, the Fisher test rejects decisively (p-value $\approx 0$).

## Comparing Estimators

Under a CRD with predictive covariates, ANCOVA should achieve substantially
smaller standard errors than DM while both remain unbiased:

```{r comparison}
results <- data.frame(
  Estimator = c("DM", "Strata (median split)", "ANCOVA"),
  ATE = NA, SE = NA
)

r <- est_via_dm(exp_obj)
results[1, c("ATE", "SE")] <- c(r$ate, r$se)

r <- est_via_strata(exp_obj, strata = strata)
results[2, c("ATE", "SE")] <- c(r$ate, r$se)

r <- est_via_ancova(exp_obj)
results[3, c("ATE", "SE")] <- c(r$ate, r$se)

knitr::kable(results, digits = 4,
             caption = "Estimator comparison (N=10000, tau=10)")
```

## Summary

- **Design objects** (`crd()`, `bernoulli()`) generate treatment assignments via `draw()`.
  CRD with `balance = TRUE` ensures covariate balance through rerandomization.
- **DM** is the natural estimator under randomization but ignores covariate information.
- **Stratified estimation** improves precision when strata are informative.
- **ANCOVA** has the largest efficiency gain when covariates predict the outcome.
- **Fisher test** gives exact p-values under the sharp null, without asymptotic
  approximations.
