---
title: "Observational Study Estimators"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Observational Study Estimators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(CausalModel)
set.seed(2024)
```

## Overview

This vignette mirrors the Python `observational.ipynb` test notebook, demonstrating
all observational-study estimators in CausalModel and validating their asymptotic
properties via Monte Carlo simulation.

## Basic Usage

### Binary Treatment

Generate data with known $\tau = 10$ and apply each estimator:

```{r basic-usage}
data <- generate_data(N = 10000, k = 5, tau = 10)
obs <- observational(data$Y, data$Z, data$X)

est_via_ols(obs)
est_via_ipw(obs)
est_via_aipw(obs)
est_via_matching(obs, num_matches = 10, num_matches_for_var = 20)
est_via_matching(obs, num_matches = 10, num_matches_for_var = 20, bias_adj = TRUE)
```

### Continuous Treatment (DML)

```{r basic-dml}
data_c <- generate_data_continuous(N = 10000, k = 5, tau = 10)
obs_c <- observational(data_c$Y, data_c$Z, data_c$X)
est_via_dml(obs_c)
```

## Simulation 1: IPW

We run 500 replications of IPW estimation to check that the studentized statistic
$T = (\hat{\tau} - \tau) / \widehat{SE}$ is approximately $N(0,1)$.

```{r sim-ipw, results="hide"}
n_reps <- 500
tau <- 10

ipw_results <- matrix(NA, nrow = n_reps, ncol = 2,
                       dimnames = list(NULL, c("ate", "se")))

for (i in seq_len(n_reps)) {
  data <- generate_data(N = 1000, k = 2, tau = tau)
  obs <- observational(data$Y, data$Z, data$X)
  r <- est_via_ipw(obs)
  ipw_results[i, ] <- c(r$ate, r$se)
}
```

```{r sim-ipw-hist, fig.height=4}
t_ipw <- (ipw_results[, "ate"] - tau) / ipw_results[, "se"]
hist(t_ipw, breaks = 30, freq = FALSE, col = "lightblue",
     main = "IPW: Studentized Statistic vs N(0,1)",
     xlab = "t", xlim = c(-4, 4))
curve(dnorm(x), add = TRUE, col = "red", lwd = 2)
abline(v = 0, lty = 2)
```

```{r sim-ipw-table}
ates <- ipw_results[, "ate"]
ses <- ipw_results[, "se"]
ci_lo <- ates - 1.96 * ses
ci_hi <- ates + 1.96 * ses
knitr::kable(data.frame(
  Bias = mean(ates) - tau,
  RMSE = sqrt(mean((ates - tau)^2)),
  Mean_SE = mean(ses),
  Emp_SE = sd(ates),
  Coverage = mean(ci_lo <= tau & tau <= ci_hi)
), digits = 3, caption = "IPW (N=1000, 500 reps)")
```

## Simulation 2: AIPW

```{r sim-aipw, results="hide"}
aipw_results <- matrix(NA, nrow = n_reps, ncol = 2,
                        dimnames = list(NULL, c("ate", "se")))

for (i in seq_len(n_reps)) {
  data <- generate_data(N = 1000, k = 2, tau = tau)
  obs <- observational(data$Y, data$Z, data$X)
  r <- est_via_aipw(obs)
  aipw_results[i, ] <- c(r$ate, r$se)
}
```

```{r sim-aipw-hist, fig.height=4}
t_aipw <- (aipw_results[, "ate"] - tau) / aipw_results[, "se"]
hist(t_aipw, breaks = 30, freq = FALSE, col = "lightblue",
     main = "AIPW: Studentized Statistic vs N(0,1)",
     xlab = "t", xlim = c(-4, 4))
curve(dnorm(x), add = TRUE, col = "red", lwd = 2)
abline(v = 0, lty = 2)
```

```{r sim-aipw-table}
ates <- aipw_results[, "ate"]
ses <- aipw_results[, "se"]
ci_lo <- ates - 1.96 * ses
ci_hi <- ates + 1.96 * ses
knitr::kable(data.frame(
  Bias = mean(ates) - tau,
  RMSE = sqrt(mean((ates - tau)^2)),
  Mean_SE = mean(ses),
  Emp_SE = sd(ates),
  Coverage = mean(ci_lo <= tau & tau <= ci_hi)
), digits = 3, caption = "AIPW (N=1000, 500 reps)")
```

## Simulation 3: Matching

Nearest-neighbor matching with bias adjustment.

```{r sim-matching, results="hide"}
match_results <- matrix(NA, nrow = n_reps, ncol = 2,
                         dimnames = list(NULL, c("ate", "se")))

for (i in seq_len(n_reps)) {
  data <- generate_data(N = 1000, k = 2, tau = tau)
  obs <- observational(data$Y, data$Z, data$X)
  r <- est_via_matching(obs, num_matches = 1, bias_adj = TRUE)
  match_results[i, ] <- c(r$ate, r$se)
}
```

```{r sim-matching-hist, fig.height=4}
t_match <- (match_results[, "ate"] - tau) / match_results[, "se"]
hist(t_match, breaks = 30, freq = FALSE, col = "lightblue",
     main = "Matching: Studentized Statistic vs N(0,1)",
     xlab = "t", xlim = c(-4, 4))
curve(dnorm(x), add = TRUE, col = "red", lwd = 2)
abline(v = 0, lty = 2)
```

```{r sim-matching-table}
ates <- match_results[, "ate"]
ses <- match_results[, "se"]
ci_lo <- ates - 1.96 * ses
ci_hi <- ates + 1.96 * ses
knitr::kable(data.frame(
  Bias = mean(ates) - tau,
  RMSE = sqrt(mean((ates - tau)^2)),
  Mean_SE = mean(ses),
  Emp_SE = sd(ates),
  Coverage = mean(ci_lo <= tau & tau <= ci_hi)
), digits = 3, caption = "Matching (N=1000, 500 reps)")
```

## Simulation 4: DML (Continuous Treatment)

Double machine learning with 2-fold cross-fitting for continuous treatment.

```{r sim-dml, results="hide"}
dml_results <- matrix(NA, nrow = n_reps, ncol = 2,
                       dimnames = list(NULL, c("ate", "se")))

for (i in seq_len(n_reps)) {
  data <- generate_data_continuous(N = 1000, k = 2, tau = tau)
  obs <- observational(data$Y, data$Z, data$X)
  r <- est_via_dml(obs, k_folds = 2)
  dml_results[i, ] <- c(r$ate, r$se)
}
```

```{r sim-dml-hist, fig.height=4}
t_dml <- (dml_results[, "ate"] - tau) / dml_results[, "se"]
hist(t_dml, breaks = 30, freq = FALSE, col = "lightblue",
     main = "DML: Studentized Statistic vs N(0,1)",
     xlab = "t", xlim = c(-4, 4))
curve(dnorm(x), add = TRUE, col = "red", lwd = 2)
abline(v = 0, lty = 2)
```

```{r sim-dml-table}
ates <- dml_results[, "ate"]
ses <- dml_results[, "se"]
ci_lo <- ates - 1.96 * ses
ci_hi <- ates + 1.96 * ses
knitr::kable(data.frame(
  Bias = mean(ates) - tau,
  RMSE = sqrt(mean((ates - tau)^2)),
  Mean_SE = mean(ses),
  Emp_SE = sd(ates),
  Coverage = mean(ci_lo <= tau & tau <= ci_hi)
), digits = 3, caption = "DML (N=1000, 500 reps)")
```

## All Histograms

Summary panel of all four estimators:

```{r all-hist, fig.height=4, fig.width=10}
par(mfrow = c(1, 4), mar = c(3, 3, 2, 1))
for (info in list(
  list(t_ipw, "IPW"), list(t_aipw, "AIPW"),
  list(t_match, "Matching"), list(t_dml, "DML"))) {
  hist(info[[1]], breaks = 30, freq = FALSE, col = "lightblue",
       main = info[[2]], xlab = "t", xlim = c(-4, 4))
  curve(dnorm(x), add = TRUE, col = "red", lwd = 2)
  abline(v = 0, lty = 2)
}
```

## Summary

All four estimators produce studentized statistics that are approximately
$N(0,1)$ under correct specification, confirming:

- Unbiasedness of the ATE estimates
- Consistency of the SE estimates
- Approximately 95% coverage of the confidence intervals
