---
title: "Getting Started with CausalModel"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with CausalModel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(CausalModel)
set.seed(42)
```

## Overview

CausalModel collects estimators for three settings:

1. **Observational studies** — OLS, IPW, AIPW, matching, DML
2. **Randomized experiments** — difference-in-means, stratified, ANCOVA, Fisher test
3. **Network interference** — clustered IPW and AIPW

This vignette shows the basic call pattern for each setting.

## Observational Study

### Generate synthetic data

The `generate_data()` function creates (Y, Z, X) with a known treatment effect `tau`.
Treatment assignment depends on covariates through a logistic propensity model, so
naive comparisons are confounded.

```{r obs-data}
data <- generate_data(N = 2000, k = 2, tau = 10)
obs <- observational(data$Y, data$Z, data$X)
```

### OLS

Linear regression of Y on (Z, X) with HC0 robust standard errors:

```{r ols}
est_via_ols(obs)
```

### IPW (Inverse Probability Weighting)

Reweights outcomes by inverse propensity scores. Uses logistic regression by default:

```{r ipw}
est_via_ipw(obs)
```

### AIPW (Doubly Robust)

Combines outcome modeling and propensity weighting — consistent if *either* model is
correctly specified:

```{r aipw}
est_via_aipw(obs)
```

### Matching

Nearest-neighbor matching with `M` matches per unit. Variance estimation follows
Abadie & Imbens (2006):

```{r matching}
est_via_matching(obs, num_matches = 3)
```

Bias-adjusted matching corrects for residual imbalance in matched covariates:

```{r matching-ba}
est_via_matching(obs, num_matches = 3, bias_adj = TRUE)
```

### DML (Double Machine Learning)

For continuous treatments, DML with cross-fitting estimates the linear treatment effect.
First, generate continuous treatment data:

```{r dml}
data_c <- generate_data_continuous(N = 2000, k = 2, tau = 10)
obs_c <- observational(data_c$Y, data_c$Z, data_c$X)
est_via_dml(obs_c)
```

### Default estimator

The `estimate()` generic dispatches to AIPW for binary treatment and DML for continuous:

```{r estimate-default}
estimate(obs)    # calls est_via_aipw
estimate(obs_c)  # calls est_via_dml
```

## Randomized Experiment

In a completely randomized design, treatment is independent of covariates.
We simulate this directly:

```{r exp-data}
N <- 1000
X <- matrix(rnorm(N * 2), ncol = 2)
tau <- 5
Z <- sample(c(rep(1, N/2), rep(0, N/2)))
Y <- tau * Z + X %*% c(1, -1) + rnorm(N)
exp_obj <- experimental(Y, Z, X)
```

### Difference-in-Means

```{r dm}
est_via_dm(exp_obj)
```

### ANCOVA

Adjusts for covariates via Y ~ 1 + Z + X + Z*X with HC0 robust SEs.
Achieves efficiency gains when covariates predict the outcome:

```{r ancova}
est_via_ancova(exp_obj)
```

### Stratified Estimator

Weighted average of within-stratum difference-in-means:

```{r strata}
strata <- ifelse(X[,1] > 0, 1, 0)
est_via_strata(exp_obj, strata = strata)
```

### Fisher Randomization Test

Tests the sharp null of no treatment effect by comparing the observed DM statistic
to its randomization distribution:

```{r fisher}
pval <- test_via_fisher(exp_obj, n_draws = 500)
cat("Fisher p-value:", pval, "\n")
```

## Network Interference

When units within clusters can affect each other's outcomes, standard estimators
are biased. `clustered()` implements the framework of Qu et al. (2021).

### Generate clustered data

```{r int-data}
cdata <- generate_fixed_cluster(
  clusters = 200,
  group_struct = c(2, 3),
  tau = 1,
  gamma = c(0.5, 0.1)
)
```

### Clustered AIPW

```{r int-aipw, results="hide"}
cl_obj <- clustered(
  Y = cdata$Y,
  Z = cdata$Z,
  X = cdata$X,
  cluster_labels = cdata$cluster_labels,
  group_labels = cdata$group_labels,
  ingroup_labels = cdata$ingroup_labels,
  n_matches = 50
)
result <- est_via_aipw(cl_obj)
```

The result is a list of group-specific estimates. Each element contains `beta_g`
(treatment effects at each neighborhood configuration) and `se` (standard errors):

```{r int-result}
for (j in seq_along(result)) {
  cat(sprintf("Group %d:\n", j - 1))
  valid <- !is.nan(result[[j]]$beta_g) & result[[j]]$se > 0
  df <- data.frame(
    g_index = which(valid) - 1,
    beta_g = round(result[[j]]$beta_g[valid], 3),
    se = round(result[[j]]$se[valid], 3)
  )
  print(df, row.names = FALSE)
  cat("\n")
}
```

### Clustered IPW

```{r int-ipw, results="hide"}
result_ipw <- est_via_ipw(cl_obj)
```

```{r int-ipw-show}
cat("Clustered IPW beta_g (group 0):\n")
valid <- !is.nan(result_ipw[[1]]$beta_g) & result_ipw[[1]]$se > 0
round(result_ipw[[1]]$beta_g[valid], 3)
```

## Custom Models

You can plug in custom ML models by providing a list with `fit()` and `predict()`
(or `predict_proba()` for classifiers).

### Random forest AIPW

```{r rf-aipw, eval=requireNamespace("randomForest", quietly=TRUE)}
rf_out <- cm_random_forest_regressor(ntree = 100)
rf_ps <- cm_random_forest_classifier(ntree = 100)
est_via_aipw(obs, outcome_model = rf_out, propensity_model = rf_ps)
```

### Custom model protocol

Any model wrapper that follows this protocol works:

```{r custom-model}
my_ridge <- list(
  fit = function(X, y) {
    X <- as.matrix(X)
    lambda <- 1.0
    p <- ncol(X)
    beta <- solve(crossprod(X) + lambda * diag(p), crossprod(X, y))
    list(beta = beta)
  },
  predict = function(model, X) {
    as.numeric(as.matrix(X) %*% model$beta)
  }
)

est_via_aipw(obs, outcome_model = my_ridge)
```
